Higher-Order unification for free!

### 2024-04-13 paper/main.tex
sec 3.1 with spellchecker

We simulate each run in \Fo{} with a run in \Ho as follows:

%Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\hoUnifPb}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
  \runH{\foUnifPb}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t,m,l) | s \in \foUnifPb, \C{s}{t}{m}{l} \}\vspace{2pt}\\
  \qquad \runHD{\hoUnifPb}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
\end{array}
$$

@@

Since each \Ho unification variables occurs together with a scope,
its assignment needs to be abstracted over it to enable the
instantiation of the same assignment to different scopes.
This is expressed by the \elpiIn{inctx} container, and in particular
its \elpiIn{abs} binding constructor.

@@

\noindent
When a single \elpiIn{fvariable} occurs multiple times with different numbers
of arguments the compiler generates multiple mappings for it, on a first
approximation, and then makes the mapping bijective by introducing
\linketa; this detail is discussed in section \ref{sec:invariant1}.

%Applying the substitution corresponds to dereferencing a term with respect to
%the memory.
%  To ease the comparison we split \Fo{} dereferencing into a
% \elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
% to memory cells that are set with their values, and has a corresponding
% operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
% has no corresponding operation in \Ho, and only ensures that
% terms of the form <<\elpiIn{fapp[fapp L1|L2]}>> are replaced by
% <<\elpiIn{fapp L3}>> where \elpiIn{L3} is the concatenation of \elpiIn{L1}
% and \elpiIn{L2}. The reasons for this asymmetry is
% that an \elpiIn{fapp} node with a flexible head is always mapped
% to a \elpiIn{uva} (as per \cref{sec:simulation,sec:beta}),
% preventing nested applications to materialize.
% \input{code/fderef}
It is worth looking at the code of \elpiIn{deref} that
applies the substitution to a \Ho{} term. Remark how assignments are moved
to the current scope, i.e. renaming the \elpiIn{abs}-bound variables
with the names in the scope of the unification variable occurrence.

@@

\noindent
Note that move strongly relies on invariant \ref{inv:uvaarity}: the length
of the arguments of all occurrences of a unification are the same. Hence
they have the same simple type for the meta-level and hence the number of
\elpiIn{abs} nodes in the assignment matches that length.
In turn this grants that \elpiIn{move} never fails.

>>>>

We simulate each run in \Fo{} with a run in \Ho as follows:
%
%Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\hoUnifPb}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
  \runH{\foUnifPb}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t,m,l) | s \in \foUnifPb, \C{s}{t}{m}{l} \}\vspace{2pt}\\
  \qquad \runHD{\hoUnifPb}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
\end{array}
$$

@@

Since each \Ho unification variable occurs together with a scope,
its assignment needs to be abstracted over it to enable the
instantiation of the same assignment to different scopes.
This is expressed by the \elpiIn{inctx} container, and in particular
its \elpiIn{abs} binding constructor.

@@

\noindent
When a single \elpiIn{fvariable} occurs multiple times with different numbers
of arguments, the compiler generates multiple mappings for it, on a first
approximation, and then ensures the mapping are bijective by introducing
\linketa; this detail is discussed in section \ref{sec:invariant1}.

%Applying the substitution corresponds to dereferencing a term with respect to
%the memory.
%  To ease the comparison we split \Fo{} dereferencing into a
% \elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
% to memory cells that are set with their values, and has a corresponding
% operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
% has no corresponding operation in \Ho, and only ensures that
% terms of the form <<\elpiIn{fapp[fapp L1|L2]}>> are replaced by
% <<\elpiIn{fapp L3}>> where \elpiIn{L3} is the concatenation of \elpiIn{L1}
% and \elpiIn{L2}. The reasons for this asymmetry is
% that an \elpiIn{fapp} node with a flexible head is always mapped
% to a \elpiIn{uva} (as per \cref{sec:simulation,sec:beta}),
% preventing nested applications to materialize.
% \input{code/fderef}
It is worth examining the code of \elpiIn{deref}, which
applies the substitution to a \Ho{} term. Notice how assignments are moved
to the current scope, i.e. the \elpiIn{abs}-bound variables are renamed
with the names in the scope of the unification variable occurrence.

@@

\noindent
Note that \elpiIn{move} strongly relies on invariant \ref{inv:uvaarity}: the length
of the arguments of all occurrences of a unification variable is the same. Hence,
they have the same simple type for the meta-level, and therefore the number of
\elpiIn{abs} nodes in the assignment matches that length.
This guarantees that \elpiIn{move} never fails.
